Concurrent Programming with CXL

### 2020-06-27 book/paper.tex
Now fonts checked through Chapter 12

A process that wants to read first waits until there are no writers:
\texttt{nwriters == 0}.  Then it increments the number of readers.
Similarly, a
process that wants to write waits until there are no readers \emph{or} writers.
Then it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$,
\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathit{nwriters} \ge n$,
\item $(\mathit{nreaders} \ge 0 \land \mathit{nwriters} = 0) \lor
    (\mathit{nreaders} = 0 \land 0 \le \mathit{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the reader/writer
specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.  Better still: if any reader is in the
critical section, any other reader is also able to enter.

@@

Figure~\ref{fig:rw2lock} presents a reader/writer lock implementation
that does not busy-wait.
You can test the code by running \texttt{cxl -m RW=RWlock RWtest}.
It uses two ordinary locks: \texttt{rwlock} is used by both readers and writers,
while \texttt{rlock} is only used by readers.
\texttt{rwlock} is held either when readers are at label \texttt{@rcs}
or when a writer is at label \texttt{@wcs}.
\texttt{rlock} is used to protect the \texttt{nreaders} variable that
counts the number of readers in the critical section.

@@

A writer simply acquires \texttt{rwlock} to enter the critical section
and releases it to exit.  The \emph{first} reader to enter the critical
section acquires \texttt{rwlock} and the \emph{last} reader to exit
the critical section releases \texttt{rwlock}.
The implementation satisfies progress: if no process is in the critical
section than any process enter.

It is instructive to see what happens when a writer is in the critical
section and two readers try to enter.  The first reader successfully
acquires \texttt{rlock} and but blocks when trying to acquire
\texttt{rwlock}, which is held by the writer.  The second reader blocks
trying acquire \texttt{rlock} because it is held by the first reader.
When the writer leaves the critical section, the first reader acquires
\texttt{rwlock}, sets \texttt{nreaders} to 1, and releases \texttt{rlock}.
\texttt{rwlock} is still held.
Then the second reader acquires \texttt{rlock} and, assuming the first
reader is still in the critical section, increments \texttt{nreaders} to 2
and enters the critical section \emph{without} acquiring \texttt{rwlock}.
\texttt{rwlock} is essentially jointly held by both readers.
It does not matter in which order they leave: the second will release
\texttt{rwlock}.

@@

\section*{Exercices}
\begin{problems}
\item Add an assertion that demonstrates that at most three
Italian philosophers are eating at the same time?
\item Implement a test program that checks to see that
the solution allows more than one Italian philosopher to eat at the same time?
\item \label{ex:qsort}
Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in  shared variable \texttt{a}.
Another shared variable, \texttt{todo}, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
that should replace the \texttt{qsort} code.  Each worker loops until \texttt{todo}
is empty and sorts the ranges that it finds until then.  The \texttt{main}
process needs to wait until all workers are done.  Use a semaphore for this:
the \texttt{main} process should use \texttt{P} to wait for each
worker to terminate.  After that it should check that the array is sorted and
that it is a permutation of the original array.
\end{problems}

@@

To illustrate how Mesa condition variables are used in practice, we demonstrate using,
once again, an implementation of reader/writer locks.
Figure~\ref{fig:RWcv} shows the code.  \texttt{rwlock} is the global lock or mutex for the
reader/writer lock.  There are two condition variables: readers wait on \texttt{rcond} and
writers wait on \texttt{wcond}.  The implementation also keeps track of the number of
readers and writers in the critical section.

@@

Clarifications:
\begin{itemize}
\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
The top value must be an address value.  The remaining values are keys into dictionaries.
The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
\item Even though CXL code does not allow taking addresses of process variables, both
shared and process variables can have addresses.
\item \texttt{Load} and \texttt{LoadVar} instructions have an optional
variable name: if omitted the top of the stack must contain the address of
the variable.
\item \texttt{Store} and \texttt{StoreVar} instructions have an optional
variable name.  In both cases the value to be assigned is on the top
of the stack.  If the name is omitted, the address is underneath that
value on the stack.
\item The effect of the \texttt{Apply} instructions depends much on $m$.
If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
and pushes the value.
If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
pushing the current program counter and setting the program counter to
$m$.  $m$ is supposed to leave the result on the stack.
\item The \texttt{Frame} instruction pushes the value of the process
register (\emph{i.e.}, the values of the process variables) onto the
stack.  It initializes the \texttt{result} variable to the empty dictionary.
The \texttt{Return} instruction restores the process register by popping
its value of the stack.
\item All method calls have exactly one argument, although it sometimes
appears otherwise:
\begin{itemize}
\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
\item \texttt{m(a)} invokes method \texttt{m} with argument \texttt{a};
\item \texttt{m(a, b, c)} invokes method \texttt{m} with tuple \texttt{(a, b, c)} as argument.
\end{itemize}
The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
\item Every \texttt{Stop} instruction must immediately be followed by a 
\texttt{Continue} instruction.
\end{itemize}

>>>>

A process that wants to read first waits until there are no writers:
\textit{nwriters} == 0.  Then it increments the number of readers.
Similarly, a
process that wants to write waits until there are no readers \emph{or} writers.
Then it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathit{nreaders} \ge n$,
\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathit{nwriters} \ge n$,
\item $(\mathit{nreaders} \ge 0 \land \mathit{nwriters} = 0) \lor
    (\mathit{nreaders} = 0 \land 0 \le \mathit{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the reader/writer
specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.  Better still: if any reader is in the
critical section, any other reader is also able to enter.

@@

Figure~\ref{fig:rw2lock} presents a reader/writer lock implementation
that does not busy-wait.
You can test the code by running \texttt{cxl -m RW=RWlock RWtest}.
It uses two ordinary locks: \textit{rwlock} is used by both readers and writers,
while \textit{rlock} is only used by readers.
\textit{rwlock} is held either when readers are at label \texttt{@rcs}
or when a writer is at label \texttt{@wcs}.
\textit{rlock} is used to protect the \textit{nreaders} variable that
counts the number of readers in the critical section.

@@

A writer simply acquires \textit{rwlock} to enter the critical section
and releases it to exit.  The \emph{first} reader to enter the critical
section acquires \textit{rwlock} and the \emph{last} reader to exit
the critical section releases \textit{rwlock}.
The implementation satisfies progress: if no process is in the critical
section than any process enter.

It is instructive to see what happens when a writer is in the critical
section and two readers try to enter.  The first reader successfully
acquires \textit{rlock} and but blocks when trying to acquire
\textit{rwlock}, which is held by the writer.  The second reader blocks
trying acquire \textit{rlock} because it is held by the first reader.
When the writer leaves the critical section, the first reader acquires
\textit{rwlock}, sets \textit{nreaders} to 1, and releases \textit{rlock}.
\textit{rwlock} is still held.
Then the second reader acquires \textit{rlock} and, assuming the first
reader is still in the critical section, increments \textit{nreaders} to 2
and enters the critical section \emph{without} acquiring \textit{rwlock}.
\textit{rwlock} is essentially jointly held by both readers.
It does not matter in which order they leave: the second will release
\textit{rwlock}.

@@

\section*{Exercices}
\begin{problems}
\item Add an assertion that demonstrates that at most three
Italian philosophers are eating at the same time?
\item Implement a test program that checks to see that
the solution allows more than one Italian philosopher to eat at the same time?
\item \label{ex:qsort}
Figure~\ref{fig:qsort} shows an iterative implementation of the Qsort
algorithm, and Figure~\ref{fig:qsorttest} an accompanying test program.
The array to be sorted is stored in  shared variable $a$.
Another shared variable, \textit{todo}, contains the ranges of the
array that need to be sorted (initially the entire array).
Re-using as much of this code as you can, implement a parallel version of
this.  You should not have to change the methods \texttt{swap}, \texttt{partition},
or \texttt{sortrange} for this.  Create \texttt{NWORKERS} ``worker processes''
that should replace the \texttt{qsort} code.  Each worker loops until \textit{todo}
is empty and sorts the ranges that it finds until then.  The \texttt{main}
process needs to wait until all workers are done.  Use a semaphore for this:
the \texttt{main} process should use \texttt{P} to wait for each
worker to terminate.  After that it should check that the array is sorted and
that it is a permutation of the original array.
\end{problems}

@@

To illustrate how Mesa condition variables are used in practice, we demonstrate using,
once again, an implementation of reader/writer locks.
Figure~\ref{fig:RWcv} shows the code.  \textit{rwlock} is the global lock or mutex for the
reader/writer lock.  There are two condition variables: readers wait on \textit{rcond} and
writers wait on \textit{wcond}.  The implementation also keeps track of the number of
readers and writers in the critical section.

@@

Clarifications:
\begin{itemize}
\item The \texttt{Address} instruction expects $n$ values $v_1, ..., v_n$ on the stack.
The top value must be an address value.  The remaining values are keys into dictionaries.
The instruction then computes the address of \texttt{\^}$v_1~v_2~...~v_n$.
\item Even though CXL code does not allow taking addresses of process variables, both
shared and process variables can have addresses.
\item \texttt{Load} and \texttt{LoadVar} instructions have an optional
variable name: if omitted the top of the stack must contain the address of
the variable.
\item \texttt{Store} and \texttt{StoreVar} instructions have an optional
variable name.  In both cases the value to be assigned is on the top
of the stack.  If the name is omitted, the address is underneath that
value on the stack.
\item The effect of the \texttt{Apply} instructions depends much on $m$.
If $m$ is a dictionary, then \texttt{Apply} finds $i$ in the dictionary
and pushes the value.
If $m$ is a program counter, then \texttt{Apply} invokes method $m$ by
pushing the current program counter and setting the program counter to
$m$.  $m$ is supposed to leave the result on the stack.
\item The \texttt{Frame} instruction pushes the value of the process
register (\emph{i.e.}, the values of the process variables) onto the
stack.  It initializes the \texttt{result} variable to the empty dictionary.
The \texttt{Return} instruction restores the process register by popping
its value of the stack.
\item All method calls have exactly one argument, although it sometimes
appears otherwise:
\begin{itemize}
\item \texttt{m()} invokes method \texttt{m} with the empty dictionary \texttt{()} as argument;
\item \texttt{m($a$)} invokes method \texttt{m} with argument $a$;
\item \texttt{m($a$, $b$, $c$)} invokes method \texttt{m} with tuple \texttt{($a$, $b$, $c$)} as argument.
\end{itemize}
The \texttt{Frame} instruction unpacks the argument to the method and places them into process variables by the given names.
\item Every \texttt{Stop} instruction must immediately be followed by a 
\texttt{Continue} instruction.
\end{itemize}
